---
title: "Montana SNOTEL Snowpack Dashboard"
author: "William Cook & Jacob Lindsey"
format: dashboard
---

```{r}
#| label: setup
#| message: false
#| warning: false

library(tidyverse)
library(lubridate)
library(knitr)
library(plotly)
library(sf)

# Import CSV
snotel = read_csv("mt_snotel_month.csv")
```

```{r}
#| label: data-cleaning

# Select & rename columns of interest
snotel = snotel |>
  select(!median_snow_depth_1991_2020_in) |>
  rename(c(lat = latitude, 
           long = longitude, 
           elev = elevation_ft,
           swe = swe_in, 
           snow_depth = snow_depth_in, 
           median_swe_1991_2020 = median_swe_1991_2020_in)
         )

# Create and populate drainage basin column
snotel = snotel |>
  mutate(
    basin = case_when(
      str_detect(huc6, "100200") ~ "Missouri Headwaters",
      str_detect(huc6, "100302") ~ "Marias",
      str_detect(huc6, "170101") ~ "Kootenai",
      str_detect(huc6, "170102") ~ "Pend Oreille",
      str_detect(huc6, "100301") ~ "Upper Missouri",
      str_detect(huc6, "100700") ~ "Upper Yellowstone",
      str_detect(huc6, "100401") ~ "Fort Peck Lake",
      str_detect(huc6, "100402") ~ "Musselshell",
      str_detect(huc6, "90400") ~ "Upper South Saskatchewan River",
      str_detect(huc6, "100500") ~ "Milk",
      TRUE ~ NA_character_)
    ) |>
  relocate(huc6, .after = basin)

# Create and populate date, month, and year columns
snotel_clean = snotel |>
  mutate(date = ymd(date)) |>
  mutate(
    month_num = month(date),
    month = month(date, label = TRUE, abbr = TRUE),
    month = factor(
      month,
      levels = c("Oct", "Nov", "Dec", "Jan", 
                 "Feb", "Mar", "Apr", "May", 
                 "Jun", "Jul", "Aug", "Sep")
    ),
    cal_year = year(date),
    water_year = if_else(month_num >= 10, 
                         cal_year + 1, 
                         cal_year)
  ) |>
  relocate(month, .after = month_year) |>
  relocate(month_num, .after = month) |>
  relocate(cal_year, .after = month_num) |>
  relocate(water_year, .after = cal_year)

# Create and populate % median SWE column
snotel_clean = snotel_clean |>
  mutate(
    pct_median_swe = if_else(
      median_swe_1991_2020 > 0,
      100 * swe / median_swe_1991_2020,
      NA_real_)
    )
  
```


# About the Data

## Row {height="527px"}

### Column
This project explores snowpack trends across the 10 drainage basins in Montana between 2015 and 2025. Our focus is on snow water equivalent (SWE), which is a measure of the amount of water contained within the snowpack. Because snowpack is a major source of water for rivers, agriculture, recreation, population centers, and many other parts of life in Montana, comparing SWE across basins allows us to investigate differences in seasonal water availability.

The Natural Resource Conservation Service (NRCS), a department of the U.S. Department of Agriculture, runs and maintains over 850 snow telemetry (SNOTEL) sites across the western U.S. and Alaska. These sites are automated monitoring stations that collect data on snowpack and climate conditions in mountain areas. SNOTEL sites commonly collect data on SWE, snow depth, precipitation, and air temperature, as well as other data related to snow and weather conditions.

Using the NRCS Air & Water Database Report Generator, we manually selected all 96 SNOTEL sites located in Montana and included the variables date, station name, station ID, latitude, longitude, elevation, SWE, snow depth, HUC6 code, and median SWE from 1991–2020. The median SWE variable represents the typical SWE value for each site and date based on the 1991–2020 reference period, which allows current observations to be compared to longer-term normal snowpack conditions.

### Column {width="750px"}
```{r}
#| label: basin_map
#| echo: false
#| warning: false
#| message: false

Sys.setenv(SHAPE_RESTORE_SHX = "YES")
library(sf)
library(maps)

# Read basin shapefile
basins_sf <- st_read("WBDHU6.shp", quiet=TRUE)

# Lookup table for basins used in project
basin_lookup <- tibble(
  huc6 = c("100200", "100302", "170101", "170102", "100301",
           "100700", "100401", "100402", "090400", "100500"),
  basin = c("Missouri Headwaters", "Marias", "Kootenai", "Pend Oreille",
            "Upper Missouri", "Upper Yellowstone", "Fort Peck Lake",
            "Musselshell", "Upper South Saskatchewan River", "Milk")
  )

# Keep only project basins
basins_sf <- basins_sf |>
  mutate(huc6 = as.character(huc6)) |>
  left_join(basin_lookup, by = "huc6") |>
  filter(!is.na(basin)) |>
  st_transform(4326)

# Montana state polygon
montana <- map("state", region = "montana", fill = TRUE, plot = FALSE) |>
  st_as_sf()

# Create one point per SNOTEL site
sites_sf <- snotel_clean |>
  distinct(lat, long, basin) |>
  filter(!is.na(lat), !is.na(long), !is.na(basin)) |>
  st_as_sf(coords = c("long", "lat"), crs = 4326, remove = FALSE)

# Basin label positions
basin_labels <- st_point_on_surface(basins_sf) |>
  mutate(
    short_basin = case_when(
      basin == "Upper South Saskatchewan River" ~ "Upper South Sask.",
      basin == "Missouri Headwaters" ~ "Missouri Hdwtrs",
      basin == "Fort Peck Lake" ~ "Fort Peck",
      TRUE ~ basin)
    )

# Create map of basins + snotel sites
ggplot() +
  geom_sf(data = montana, fill = "gray95", color = "black", linewidth = 0.4) +
  geom_sf(data = basins_sf, aes(fill = basin), color = "gray40", linewidth = 0.3, alpha = 0.6) +
  geom_sf(data = sites_sf, color = "black", size = 1.2, alpha = 0.8) +
  geom_sf_label(data = basin_labels, aes(label = short_basin), size = 3, label.size = 0.2, 
                fill = "white", alpha = 0.7) +
  labs(
    title = "Montana Drainage Basins and SNOTEL Site Locations",
    subtitle = "SNOTEL sites shown as black points",
    fill = "Drainage Basin"
  ) +
  theme_minimal() +
  theme(
    legend.position = "none",
    panel.grid.major = element_blank()
  )
```

## Row {height="200px"}
Observations were recorded from the 1st of each month instead of daily observations, which greatly reduced the size of our dataset and allows for stations and years to be compared directly. The term "year" refers to the water year (abreviated to "WY"), which runs from October 1 to September 30 of the following calendar year (ex: WY 2025 begins on 10/01/2024 and ends 09/30/2025). The dataset also contains six-digit Hydrologic Unit Codes (HUC6) used to identify watershed regions or basins. These codes were used to group individual SNOTEL sites into major Montana drainage basins, allowing SWE patterns to be compared by watershed

The monthly data does not capture short-term snow or melt events, so this dashboard is best used to compare broad seasonal and basin-level SWE patterns rather than day-to-day snowpack conditions.




# Seasonal SWE Patterns

## Row {height="500px"}

### Column {width="60%"}

::: {.card}
```{r}
#| label: mean-monthly-swe-plot
#| fig-height: 4.22
#| fig-width: 7

# Create avg SWE by month/basin plot
snotel_clean |>
  group_by(basin, month) |>
  summarise(mean_swe = mean(swe, na.rm = TRUE), .groups = "drop") |>
  ggplot(aes(x = month, 
             y = mean_swe, 
             color = basin, 
             group = basin)) + 
    geom_line() +
    labs(
      title = "Mean Monthly SWE by Basin (WY 2016–2025)",
      x = "Month",
      y = "Mean Snow Water Equivalent (in)") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
:::

### Column {width="40%"}

:::{.card height="460px"}
```{r}
#| label: basin-summary-table

# Create Basin-SWE summary table
snotel_clean |>
  group_by(basin) |>
    summarise(
      `Mean SWE` = round(mean(swe, na.rm = TRUE), 2),
      `Median SWE` = round(median(swe, na.rm = TRUE), 2),
      `Maximum SWE` = round(max(swe, na.rm = TRUE), 2),
      .groups = "drop") |>
    arrange(desc(`Mean SWE`)
          ) |>
    kable(
      caption = "Summary of SWE by Drainage Basin",
      col.names = c("Drainage Basin", "Mean SWE", "Median SWE", "Maximum SWE")
      )
```
:::

## Row

:::{.card height="180px"}
Across most basins, SWE follows a clear seasonal pattern. Snowpack begins building in the fall and winter, reaches its highest values in the spring, and then
declines quickly during late spring and summer. This pattern reflects Montana’s seasonal snow accumulation and melt cycle.

The summary table shows that the basins Kooteni, Pend Oreille, and Fort Peck Lake consistently have higher SWE than others. These differences are likely related to elevation,
mountain range location, and regional climate patterns. Because the dashboard averages across all stations in each basin, the results are best interpreted as
broad basin-level patterns rather than conditions at one specific SNOTEL site.
:::

# SWE Comparison

## Row {height="690px"}

### Column {width="65%"}
::: {.card}
```{r}
#| label: interactive-pct-median-swe-heatmap

# Filter/group data needed for heatmap
heatmap_data = snotel_clean |>
  filter(month %in% c("Dec", "Jan", "Feb", "Mar", "Apr", "May")) |>
  mutate(
    snow_month_num = case_when(
      month == "Dec" ~ 1,
      month == "Jan" ~ 2,
      month == "Feb" ~ 3,
      month == "Mar" ~ 4,
      month == "Apr" ~ 5,
      month == "May" ~ 6)
    ) |>
  group_by(water_year, basin, month, snow_month_num) |>
  summarise(
    mean_pct_median_swe = mean(pct_median_swe, na.rm = TRUE),
    .groups = "drop"
    ) |>
  filter(!is.na(mean_pct_median_swe))

# Create interactive heatmap plot
heatmap = plot_ly(
  data = heatmap_data,
  x = ~snow_month_num,
  y = ~basin,
  z = ~mean_pct_median_swe,
  frame = ~water_year,
  type = "heatmap",
  colors = c(
    "#b2182b",
    "#ef8a62",
    "#fddf8a",
    "#a6d96a",
    "#66bd63",
    "#1a9850"),
  zmin = 0,
  zmax = 200,
  colorbar = list(
    title = "% of<br>Median SWE",
    tickvals = c(0, 50, 75, 100, 125, 150, 200),
    ticktext = c("0%", "50%", "75%", "100%", "125%", "150%", "200%+")),
  text = ~paste0(
    "Water Year: ", water_year,
    "<br>Month: ", month,
    "<br>Basin: ", basin,
    "<br>% of Median SWE: ", 
    round(mean_pct_median_swe, 1), 
    "%"),
    hovertemplate = "%{text}<extra></extra>"
  ) |>
  layout(
    title = "Percent of Median SWE by Basin and Month",
    height = 650,
    margin = list(l = 140, r = 140, t = 30, b = 40),
    xaxis = list(
      title = "Month",
      tickmode = "array",
      tickvals = 1:6,
      ticktext = c("Dec", "Jan", "Feb", "Mar", "Apr", "May"),
      range = c(0.5, 6.5),
      fixedrange = TRUE),
    yaxis = list(
      title = "Drainage Basin",
      fixedrange = TRUE)
    ) |>
  animation_opts(
    frame = 0,
    transition = 0,
    redraw = TRUE
    ) |>
  animation_slider(
    currentvalue = list(prefix = "Water Year: ")
    )

heatmap
```
:::

### Column {width="35%"}
This heatmap compares observed SWE to the 1991–2020 median SWE during the main snowpack season. Use the water year slider to change the water year. Values 
near 100% indicate SWE close to the long-term median, values below 100% indicate below-median snowpack, and values above 100% indicate above-median snowpack.

#### Row

```{r}
#| warning: false
# Create peak swe tibble, filter and order months, and group sites by wy & basin
peak_data <- snotel_clean |>
  filter(month %in% c("Jan", "Feb", "Mar", "Apr", "May", "Jun")) |>
  group_by(water_year, basin, month) |>
  summarise(
    mean_swe = mean(swe, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(
    peak_month_num = case_when(
      month == "Jan" ~ 1,
      month == "Feb" ~ 2,
      month == "Mar" ~ 3,
      month == "Apr" ~ 4,
      month == "May" ~ 5,
      month == "Jun" ~ 6
    )
  ) |>
  group_by(water_year, basin) |>
  slice_max(order_by = mean_swe, n = 1, with_ties = FALSE) |>
  ungroup()

# Create summary of peak SWE month by basin across all years in data
peak_summary <- peak_data |>
  group_by(basin) |>
  summarise(
    median_peak_month_num = median(peak_month_num, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(
    `Peak Month` = case_when(
      median_peak_month_num == 1 ~ "Jan",
      median_peak_month_num == 2 ~ "Feb",
      median_peak_month_num == 3 ~ "Mar",
      median_peak_month_num == 4 ~ "Apr",
      median_peak_month_num == 5 ~ "May",
      median_peak_month_num == 6 ~ "Jun"
    )
  ) |>
  arrange(desc(median_peak_month_num)) |>
  select(basin, `Peak Month`)

# Run KW test to determine true differences in peak SWE month
peak_test <- kruskal.test(peak_month_num ~ basin, data = peak_data)
```

```{r}
# Create peak SWE month by basin table
peak_summary |>
  knitr::kable(
    caption = "Typical Peak Snowpack Month by Basin",
    col.names = c("Basin", "Peak Month"))
```

```{r}
#| output: false

# Run pairwise wilcoxon test for more info about basin differences
pairwise_peak <- pairwise.wilcox.test(
  x = peak_data$peak_month_num,
  g = peak_data$basin,
  p.adjust.method = "holm",
  exact = FALSE)
pairwise_peak
```

# Findings and Resources


## Row {height="240px"}

### Column {width="41%"}
In this case study we are focusing on the median SWE for each snow telemetry site in Montana from 2016 to 2025. After manually assigning each SNOTEL site in Montana to its respective drainage basin using the HUC6 code, we were able to compare basin wide SWE observations to the median SWE from 1991 to 2020 by both months and years. Doing this we are able to visualize the difference in snow water equivalent not only for each drainage basin or region in Montana, but compare the same basin to itself during different water years.

### Column {width="59%"}
A Kruskal-Wallis test found strong evidence that the typical peak snowpack month differs among basins ($\chi^2 = 49.311, df = 9, p < 0.001$). Pairwise Wilcoxon tests showed the strongest differences involved the Milk, Fort Peck Lake, and Musselshell basins, indicating that peak snowpack timing is not uniform across Montana drainage basins. The clearest pattern was that the Milk basin typically peaked earlier, in March, while Fort Peck Lake and Musselshell basins typically peaked later, in May. Most other basins most often peaked in April.
Because the dataset only includes observations from the first day of each month, these results should be interpreted as differences in peak timing at the monthly scale rather than the exact date of annual maximum snowpack.

## Row
Through this analysis, it is hard to compare actual amount of snow in one region to another, as our reference is each region's median snow water equivalent
over the last 30 years. What these graphics excel at is seeing each region's percent of median SWE for each month. We can analyze the differences throughout
the years, and even see the changes in percent throughout one year. Although we will miss smaller scale weather events such as melts or large storms because
of our monthly readings, this information still gives us great insights on historic patterns of snow melt for the state of Montana. 

SWE and snowpack have more implications than just tracking the snow; one of the larger parts of watching the seasonal snowpack is hydrological health. 
When we look at the Fort Peck Lake and Musselshell basins, we see their usual peaks at the end of the season in April and May. This important because as we move into Eastern Montana, we start to see a lot more of agricultural activity that relies on large amounts of water for irrigation and livestock livability. Having a snowpack that lasts into May and delays runoff is critical to having persistent farms that will continue to supply crops throughout the dry summer months.
This also brings out a point to mention where these basins in Eastern Montana may be underrepresented. Most of these SNOTEL sites are focused in higher elevation, wetter climates where more snow tends to fall, so there is not nearly as much data in these drier, flatter basins. 

Through this analysis, it is hard to compare actual amount of snow in one region to another, as our reference is each region's median snow water equivalent
over the last 30 years. What these graphics excel at is seeing each region's percent of median SWE for each month. We can analyze the differences throughout
the years, and even see the changes in percent throughout one year. Although we will miss smaller scale weather events such as melts or large storms because
of our monthly readings, this information still gives us great insights on historic patterns of snow melt for the state of Montana. 

## Row
**Resources**
U.S. Department of Agriculture, Natural Resources Conservation Service. (n.d.). Report Generator 2.0: Air & Water Database Report Generator. Retrieved April 1, 2026.
https://wcc.sc.egov.usda.gov/reportGenerator/

U.S. Department of Agriculture, Natural Resources Conservation Service, National Geospatial Center of Excellence, & U.S. Geological Survey. (2025, February 4). Watershed Boundary Dataset for Montana [Data set]. Montana State Library.
https://mslservices.mt.gov/Geographic_Information/Data/DataList/datalist_Details.aspx?did=%7B0077A7D3-F6C3-4D08-9837-410E79A12FCC%7D

